---
title: "Interpreting chronic AmsPAF outputs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interpreting chronic AmsPAF outputs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

This vignette covers how to interpret the values returned by `add_amspaf()`
and `time_weighted_aggregate()` in the context of an environmental
assessment. The package deliberately stops short of providing regulatory
tier breaks — those depend on the assessment context and are a consumer
decision. This document gives you the conceptual tools to make those
decisions.

## msPAF is not a single-substance trigger value

The "5%" threshold in ANZG/ANZECC guidelines (the 95% species protection
level) is derived from a **single-substance** species sensitivity
distribution (SSD). It says: "if a water body contains only this substance
at this concentration, 5% of species would be affected."

AmsPAF outputs a number on the same 0–100% scale but with a different
denominator: "given the *mixture* of substances present at observed
concentrations, what fraction of species would be affected by *the entire
mixture*?"

The implications:

1. A 5% AmsPAF and a 5% single-substance PAF do not correspond to the same
   "level of impact." A mixture of 10 substances each at their 1% PAF
   contributes a combined AmsPAF substantially larger than any single
   substance's 1%.

2. The denominator depends on which substances enter the calculation. A
   site with Cu and Zn measured will have a different AmsPAF than the same
   site with Cu, Zn, and Pb measured — even if Pb is well below its
   single-substance trigger.

3. **You cannot directly compare an AmsPAF value to an ANZG
   single-substance protection level.** They are different metrics on the
   same scale.

What you *can* do:

- Compare AmsPAF across sites to identify relative impact.
- Compare AmsPAF over time at one site to detect trends.
- Calibrate AmsPAF against biological response data (e.g.
  macroinvertebrate community state) to derive site-relevant interpretive
  thresholds.

## How the mixture is combined (and the log-normal assumption)

AmsPAF combines substances in two stages, following De Zwart & Posthuma
(2005). Substances thought to act by the same mechanism are combined within
a group by **concentration addition (CA)**; the groups are then combined
across by **independent action (IA)**. The two stages make different
assumptions about the underlying species sensitivity distributions, and it is
worth knowing which is which.

- **Independent action across groups is shape-agnostic.** IA combines groups
  as `msPAF = 1 − ∏(1 − PAF_i)`. It only needs each component's PAF, which the
  package reads off the full model-averaged SSD (all six BCANZ distributions,
  weighted) — so this step holds whatever shape the SSDs take.

- **Concentration addition within a group assumes a log-normal.** CA sums the
  toxic units (`TU_i = C_i / HC50_i`) and reads the combined fraction off a
  *single* log-normal curve whose slope is the average of the component slopes.
  This is the standard msPAF-CA method, and it is **exact when the component
  SSDs are log-normal** (or log-logistic). When the best-fitting SSD for a
  substance is some other shape, the CA step approximates it with a log-normal
  matched at its HC50 and HC05. For the metals that dominate leachate
  assessments this approximation is small — their SSDs sit close to log-normal
  — but it is a modelling choice, not an identity.

A future release will add a shape-faithful numerical CA (which makes no
log-normal assumption) as an option; until then, if a substance's SSD is
strongly non-log-normal, treat its contribution to the within-group CA as
approximate.

## Chronic vs per-sample AmsPAF

`add_amspaf()` operates on a chemistry data frame regardless of whether
each row is a single field sample or a chronic-aggregated value — it
treats every row as a snapshot of exposure and computes the AmsPAF that
snapshot implies.

For chronic interpretation (i.e. comparing AmsPAF against community state
that integrates exposure over weeks to months), there are two pathways:

**Path A — concentration averaging.** Time-aggregate chemistry first, then
compute AmsPAF on the aggregated chemistry:

```{r eval = FALSE}
chr_chem <- time_weighted_aggregate(imp, focal_dates, summary = "geom_mean")
prep_ref <- prepare_reference(chr_ref)
chr_amspaf <- add_amspaf(chr_chem, reference = prep_ref)
```

This gives "AmsPAF at the geometric-mean chronic chemistry." For sites
with stable chemistry it is acceptable, but for sites with pulsed
exposures (storm events, mobilisation pulses) it can substantially
under-state the toxic burden — see *Pulsed exposures* below.

**Path B — response averaging (recommended).** Compute per-sample AmsPAF
first, then time-aggregate the AmsPAF values:

```{r eval = FALSE}
prep_ref       <- prepare_reference(ref_chem)
ps_amspaf      <- add_amspaf(imp, reference = prep_ref)
chr_amspaf     <- time_weighted_aggregate(
  dplyr::filter(ps_amspaf, analyte == "AmsPAF"),
  focal_dates    = focal_dates,
  summary        = "arith_mean"
)
```

This time-averages the toxic-response signal directly. It is more
biologically defensible (the community integrates the effect over time,
not the concentration) and is the default in the bundled
`run_amspaf_pipeline.R` script.

### Pulsed exposures and Jensen's inequality

For a log-normal SSD with HC50 = 100 µg/L and σ_log10 = 0.5, consider
100 days of exposure with 99 days at 1 µg/L (baseline) and 1 day at
1000 µg/L (storm pulse):

| Aggregation | Concentration | PAF |
|---|---|---|
| Geometric mean concentration | 1.99 µg/L | 0.03 % |
| Arithmetic mean concentration | 11.0 µg/L | 2.7 % |
| Mean of per-sample PAFs | — | 0.91 % |
| Max concentration | 1000 µg/L | 97.7 % |

PAF at the geometric mean concentration under-states the time-averaged
PAF by an order of magnitude. This is Jensen's inequality acting on the
concave upper tail of the SSD: a brief excursion to high concentration
contributes much more to the integrated response than its concentration-
space weight suggests.

For chronic assessment of pulsed exposures, Path B (mean of per-sample
AmsPAF values) is the correct aggregation.

## Choosing `tau_days`

`time_weighted_aggregate()` uses exponential-decay temporal weighting with
parameter `tau_days` (default 90). The effective half-life is
`tau_days * log(2)` ≈ 62 days at the default.

Choose `tau_days` to match the response time of the target biology:

| Target | Typical response timescale | Suggested `tau_days` |
|---|---|---|
| Periphyton / algae | days to weeks | 14–30 |
| Macroinvertebrates (general) | weeks to months | 60–120 |
| Fish (sublethal effects) | months to years | 180–365 |
| Long-lived benthic communities | seasonal to multi-year | 365+ |

These are starting points, not prescriptions. For a calibration exercise
the right move is to test a range of `tau_days` values and pick the one
that maximises explanatory power against your biological response data —
within a range that is mechanistically defensible.

## Reference summary statistic

The Added Risk Approach (ARA) subtracts a background "reference"
concentration before evaluating the SSD, so the AmsPAF reflects only
the *increment above* what the local community is already adapted to.
The summary statistic used to define that background is chosen by the
`summary` argument to `prepare_reference()`.

The package default is `"geom_mean"`. Rationale:

- For log-normal concentration distributions (the typical case for
  aquatic chemistry), the geometric mean is the maximum-likelihood
  central tendency.
- It uses every reference observation, making it more robust to small
  reference datasets than a fixed quantile.
- It is consistent with the PICT (pollution-induced community
  tolerance) interpretation: the community has been integrated against
  the typical exposure, not the upper tail.

Alternatives:

- `"median"` — most robust to outliers; good when reference data may be
  contaminated.
- `"arith_mean"` — uses every observation but is sensitive to outliers.
- `"p80"`, `"p90"`, `"p95"` — quantiles. Use when the assessment
  framework specifies an upper-percentile reference (e.g., some ANZG
  compliance protocols).

Set `bootstrap_ci = TRUE` to compute a 95% bootstrap CI on the chosen
summary statistic per analyte. Useful when the reference dataset is
small (n < 20 per analyte) and you want to propagate that uncertainty
visually into downstream interpretation.

## Known caveats

**Mercury and methylation.** The Hg SSD bundled with this package is for
inorganic Hg. Methylmercury (MeHg), which forms in low-redox aquatic
sediments and is the form most toxic to vertebrates, is not separately
handled — ANZG does not publish a methylmercury SSD. At sites where
reductive conditions are expected (leachate plumes, anaerobic sediment),
the inorganic Hg PAF will under-state the actual biological risk by
several orders of magnitude. This is a limitation in upstream guideline
data, not in the package.

**Mode-of-action grouping.** The current metadata places all metals into
a single `ionoregulatory` group (Concentration Addition within the
group). This is the conservative choice — it produces a higher AmsPAF
than independent action across the metals — but it is mechanistically
imprecise. Different metals act on different molecular targets (Cu/Zn on
gill Na⁺/K⁺-ATPase, Pb on calcium signalling, Cd on metallothionein,
Hg on thiol groups, As on phosphate metabolism). A more accurate
treatment would use multiple MOA groups with CA within and IA across.
This requires evidence-based assignment from the ANZG/ANZECC technical
briefs, which is tracked as a future task.

**Biological validation.** The SSDs are validated against published ANZG
DGVs (see HANDOVER.md §4) but the AmsPAF as a community-state predictor
is not yet calibrated. Until macroinvertebrate calibration is complete,
AmsPAF outputs are best-effort predictions of chronic toxicity; they
have not been empirically tied to observed community impact.

**Below-detection handling in imputation.** `fit_imputation_model()` offers
three `impute_method` options for BDL cells and cross-analyte coupling:
`"rescor_mi"` (default — full residual correlation with `mi()` imputation),
`"cens"` (proper left-censoring, no coupling), and `"cens_factor"`
(left-censoring plus a shared latent factor). The default cannot use
`cens("left")` because brms does not support censored likelihoods together with
multivariate residual correlation (`rescor = TRUE`), so it imputes BDL cells via
`mi()`. In all methods an imputed BDL value is capped at its detection limit
(`bdl_cap = TRUE`); frequent capping signals tension between the modelled
chemistry and the reported limits and should be inspected. See the *Imputing
missing and below-detection chemistry* article and the `fit_imputation_model()`
docs for the trade-offs.
