---
title: "Interpreting chronic AmsPAF outputs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interpreting chronic AmsPAF outputs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

This vignette covers how to interpret the values returned by `add_amspaf()`
and `time_weighted_aggregate()` in the context of an environmental
assessment. The package deliberately stops short of providing regulatory
tier breaks — those depend on the assessment context and are a consumer
decision. This document gives you the conceptual tools to make those
decisions.

## msPAF is not a single-substance trigger value

The "5%" threshold in ANZG/ANZECC guidelines (the 95% species protection
level) is derived from a **single-substance** species sensitivity
distribution (SSD). It says: "if a water body contains only this substance
at this concentration, 5% of species would be affected."

AmsPAF outputs a number on the same 0–100% scale but with a different
denominator: "given the *mixture* of substances present at observed
concentrations, what fraction of species would be affected by *the entire
mixture*?"

The implications:

1. A 5% AmsPAF and a 5% single-substance PAF do not correspond to the same
   "level of impact." A mixture of 10 substances each at their 1% PAF
   contributes a combined AmsPAF substantially larger than any single
   substance's 1%.

2. The denominator depends on which substances enter the calculation. A
   site with Cu and Zn measured will have a different AmsPAF than the same
   site with Cu, Zn, and Pb measured — even if Pb is well below its
   single-substance trigger.

3. **You cannot directly compare an AmsPAF value to an ANZG
   single-substance protection level.** They are different metrics on the
   same scale.

What you *can* do:

- Compare AmsPAF across sites to identify relative impact.
- Compare AmsPAF over time at one site to detect trends.
- Calibrate AmsPAF against biological response data (e.g.
  macroinvertebrate community state) to derive site-relevant interpretive
  thresholds.

## How the mixture is combined (and the log-normal assumption)

AmsPAF combines substances in two stages, following De Zwart & Posthuma
(2005). Substances thought to act by the same mechanism are combined within
a group by **concentration addition (CA)**; the groups are then combined
across by **independent action (IA)**. The two stages make different
assumptions about the underlying species sensitivity distributions, and it is
worth knowing which is which.

- **Independent action across groups is shape-agnostic.** IA combines groups
  as `msPAF = 1 − ∏(1 − PAF_i)`. It only needs each component's PAF, which the
  package reads off the full model-averaged SSD (all six BCANZ distributions,
  weighted) — so this step holds whatever shape the SSDs take.

- **Concentration addition within a group assumes a log-normal.** CA sums the
  toxic units (`TU_i = C_i / HC50_i`) and reads the combined fraction off a
  *single* log-normal curve whose slope is the average of the component slopes.
  This is the standard msPAF-CA method, and it is **exact when the component
  SSDs are log-normal** (or log-logistic). When the best-fitting SSD for a
  substance is some other shape, the CA step approximates it with a log-normal
  matched at its HC50 and HC05. For the metals that dominate leachate
  assessments this approximation is small — their SSDs sit close to log-normal
  — but it is a modelling choice, not an identity.

A future release will add a shape-faithful numerical CA (which makes no
log-normal assumption) as an option; until then, if a substance's SSD is
strongly non-log-normal, treat its contribution to the within-group CA as
approximate.

## Chronic vs per-sample AmsPAF

`add_amspaf()` operates on a chemistry data frame regardless of whether
each row is a single field sample or a chronic-aggregated value — it
treats every row as a snapshot of exposure and computes the AmsPAF that
snapshot implies.

For chronic interpretation (i.e. comparing AmsPAF against community state
that integrates exposure over weeks to months), there are two pathways:

**Path A — concentration averaging.** Time-aggregate chemistry first, then
compute AmsPAF on the aggregated chemistry:

```{r eval = FALSE}
chr_chem <- time_weighted_aggregate(imp, focal_dates, summary = "geom_mean")
prep_ref <- prepare_reference(chr_ref)
chr_amspaf <- add_amspaf(chr_chem, reference = prep_ref)
```

This gives "AmsPAF at the geometric-mean chronic chemistry." For sites
with stable chemistry it is acceptable, but for sites with pulsed
exposures (storm events, mobilisation pulses) it can substantially
under-state the toxic burden — see *Pulsed exposures* below.

**Path B — response averaging (recommended).** Compute per-sample AmsPAF
first, then time-aggregate the AmsPAF values:

```{r eval = FALSE}
prep_ref       <- prepare_reference(ref_chem)
ps_amspaf      <- add_amspaf(imp, reference = prep_ref)
chr_amspaf     <- time_weighted_aggregate(
  dplyr::filter(ps_amspaf, analyte == "AmsPAF"),
  focal_dates    = focal_dates,
  summary        = "arith_mean"
)
```

This time-averages the toxic-response signal directly. It is more
biologically defensible (the community integrates the effect over time,
not the concentration) and is the default in the bundled
`run_amspaf_pipeline.R` script.

### Pulsed exposures and Jensen's inequality

For a log-normal SSD with HC50 = 100 µg/L and σ_log10 = 0.5, consider
100 days of exposure with 99 days at 1 µg/L (baseline) and 1 day at
1000 µg/L (storm pulse):

| Aggregation | Concentration | PAF |
|---|---|---|
| Geometric mean concentration | 1.99 µg/L | 0.03 % |
| Arithmetic mean concentration | 11.0 µg/L | 2.7 % |
| Mean of per-sample PAFs | — | 0.91 % |
| Max concentration | 1000 µg/L | 97.7 % |

PAF at the geometric mean concentration under-states the time-averaged
PAF by an order of magnitude. This is Jensen's inequality acting on the
concave upper tail of the SSD: a brief excursion to high concentration
contributes much more to the integrated response than its concentration-
space weight suggests.

For chronic assessment of pulsed exposures, Path B (mean of per-sample
AmsPAF values) is the correct aggregation.

## Choosing `tau`

`time_weighted_aggregate()` uses exponential-decay temporal weighting with
parameter `tau` (default `NULL` → 90 days; pass e.g. `tau = 90, tau_units = "d"`).
The effective half-life is `tau * log(2)` ≈ 62 days at the default.

Choose `tau` to match the response time of the target biology:

| Target | Typical response timescale | Suggested `tau` (days) |
|---|---|---|
| Periphyton / algae | days to weeks | 14–30 |
| Macroinvertebrates (general) | weeks to months | 60–120 |
| Fish (sublethal effects) | months to years | 180–365 |
| Long-lived benthic communities | seasonal to multi-year | 365+ |

These are starting points, not prescriptions. For a calibration exercise
the right move is to test a range of `tau` values and pick the one
that maximises explanatory power against your biological response data —
within a range that is mechanistically defensible.

## Contemporaneous (temporal) ARA reference

The default ARA path subtracts a *static* background — a geometric mean (or
another summary statistic) computed once from all available reference
observations. This is appropriate when the reference dataset is large, well
distributed across seasons and flow regimes, and the key question is long-run
chronic risk.

However, in flashy catchments or sites with strong seasonal chemistry,
subtracting a static mean can mismatch the target sample: if the target was
collected during a storm flush and the reference background used is the
dry-weather mean, the ARA will either over-correct (geogenic storm-flush metals
exceeding the dry-weather mean are erroneously attributed to site impact) or
under-correct, depending on the season.

`fit_reference_model()` addresses this by fitting **per-analyte seasonal +
hydrological models** on the reference chemistry.  Given a daily hydrology
series (rainfall → API, or stage/discharge → antecedent mean), the function
trains a GAM with a cyclic day-of-year term and two Antecedent Precipitation
Index (API) terms with different memory lengths:

```
log(ref_norm) ~ s(doy, bs="cc") + s(API_short) + s(API_long)
```

When you pass the resulting `reference_model` to `add_amspaf()`, the
*contemporaneous* reference norm is resolved per (sample × analyte) via a
three-tier ladder:

1. **Tier 1 — direct match.** A reference grab within `±match_window_days`
   days of the target sample *and* within an API-tolerance gate (hydrologically
   similar moment) is used directly.  The API gate prevents using a dry-weather
   reference observation next to a storm-event target.

2. **Tier 2 — GAM prediction.** If no matched grab exists, the per-analyte
   GAM predicts the expected reference at the target's hydrology + season.
   Only used when the model beats a null (intercept-only) baseline on AIC; the
   window lengths (short and long API memory) are auto-selected by AIC over a
   candidate grid.

3. **Tier 3 — static fallback.** If the GAM is not better than the null (e.g.,
   reference chemistry shows no seasonal or hydrological pattern — common for
   conservative analytes), the geometric mean of all reference observations is
   used, matching the static `prepare_reference()` path.

**Chronic integration.** For chronic samples produced by
`time_weighted_aggregate()`, the temporal path integrates at daily resolution
across the `window` look-back period using the same exponential-decay kernel
(`tau`).  This makes the chronic reference consistent with the chronic target.

**Quick start:**

```r
# Fetch SILO daily rainfall automatically from lat/lon
ref_model <- fit_reference_model(
  reference  = reference_chemistry,   # long-format, detected = TRUE
  latitude   = -33.87,
  longitude  = 151.21,
  conc_units = "ug/L"
)

# Pass to add_amspaf() in place of a static reference
out <- add_amspaf(target_chem, reference = ref_model, conc_units = "ug/L")

# Inspect per-cell ARA diagnostics (ref tier, floor events, etc.)
ara_summary(out)
```

Supply your own gauge record when you have stage or discharge data:

```r
stage_df  <- data.frame(date = ..., value = ...)   # daily stage in m
ref_model <- fit_reference_model(
  reference  = reference_chemistry,
  hydro      = stage_df,
  hydro_type = "stage",
  conc_units = "ug/L"
)
```

**`ara_summary()` diagnostic.** Every `add_amspaf()` call (regardless of
reference type) stores a per-(sample × analyte) diagnostic tibble as an
attribute.  `ara_summary(out)` returns it:

| column | meaning |
|---|---|
| `C_norm` | Normalised target concentration |
| `ref_norm` | Normalised reference subtracted |
| `C_adj` | `max(C_norm − ref_norm, 0)` (the floored ARA increment) |
| `C_excess` | Signed `C_norm − ref_norm` (negative = reference exceeded target) |
| `floor_fired` | `TRUE` when `C_excess < 0` — reference higher than target |
| `ref_tier` | `"matched"`, `"model"`, `"model_integrated"`, or `"static"` |

The `floor_fired` rows are the most ecologically interesting: they indicate
analytes where the reference site is *more* contaminated than the target.  This
typically happens for pH-mobile metals (Zn, Ni, Cu) when a low-pH headwater
reference releases metals that precipitate at the higher-pH target — a real
geogenic effect, not a data error.  These cells are correctly zeroed in the
ARA, but `ara_summary()` surfaces them so you can document the reason.

**Package invariant (shared catchment).** `fit_reference_model()` assumes
reference and target share the same catchment hydrology.  One hydrology series
drives the API computation for both.  This is appropriate for the canonical
leachate-assessment design (headwater reference, downstream target, same
sub-catchment) and should be stated explicitly in any report.

---

## Predicting the target site between grabs

`fit_reference_model()` predicts the *background*. `fit_target_model()` is its
counterpart for the *impacted* site: it predicts the leachate-attributable
increment

$$I(t) = C_\text{norm}(t) - \text{ref}_\text{norm}(t)$$

— the very quantity `ara_summary()` reports as `C_excess` — so that
`amspaf_daily(interpolation = "model")` can fill the days between grab samples
with a chemistry-grounded estimate instead of a forward-filled concentration.

**The crucial asymmetry: the target model is season-blind.** This is the one
place where the target and reference logic must *differ*. Natural background
chemistry genuinely follows the calendar (leaf litter, baseflow recession,
snowmelt), so the reference model legitimately carries a cyclic day-of-year
term. Site impacts do not. They are driven by **management failure** — a
leachate breach — not by the season. Heavy summer rain *enables* an existing
breach to escape; it does not *cause* one. A year of perfect management has zero
impact despite identical rainfall. Letting day-of-year predict impact would
hallucinate a breach from the calendar — for example, forecasting elevated
metals every February simply because past Februaries happened to coincide with
breaches. The target model therefore excludes day-of-year entirely. Hydrology
enters only as a *modulator* of how an already-present impact expresses itself
(first-flush mobilisation, dilution, antecedent memory), never as a generator:

$$I(t) = \beta \cdot f(\text{hydro}_t) + S(t)$$

`f(hydro)` is a thin-plate GAM on the short- and long-window antecedent indices
(no seasonal term), kept only when it beats an intercept-only null by AIC. `S(t)`
is a persistent latent state interpolated between observation anchors by a
hydrology-weighted bridge: it pinches to the observed residual on each sampling
day, and between two grabs it leans toward whichever bracketing sample's
antecedent hydrology more closely matches the day being predicted. The
**perfect-management invariant** falls out by construction — when every observed
impact is ≈ 0, the fitted response is flat and the interpolated state stays ≈ 0,
so no impact is invented on un-sampled high-rainfall days.

This is, in effect, a season-blind, hydrology-modulated variant of
**WRTDS-Kalman** (Zhang & Hirsch 2019) applied to the ARA impact residual; the
hydrological response `f(hydro)` follows **concentration–discharge** theory
(Godsey, Kirchner & Clow 2009), which is why `\beta` is left free-signed
per analyte (a first-flush metal rises with antecedent rainfall; a
dilution-controlled one falls).

```r
ref_model <- fit_reference_model(reference_chem, latitude = -33.8,
                                 longitude = 151.2, conc_units = "ug/L")

# Daily AmsPAF (leachate-attributable increment) — model interpolation + ARA
daily_ara <- amspaf_daily(
  target_chem,
  reference       = ref_model,   # subtract background (ARA on)
  reference_model = ref_model,   # season-blind impact model interpolates toxicants
  interpolation   = "model"
)

# Daily total msPAF — same interpolation, ARA off (reference = NULL)
daily_total <- amspaf_daily(
  target_chem,
  reference       = NULL,
  reference_model = ref_model,
  interpolation   = "model"
)
```

Note that `reference` and `reference_model` play distinct roles:
`reference_model` is what *interpolates* the toxicants (it is required for the
`"model"` path), while `reference` independently controls whether the background
is *subtracted*. Passing the same model to both gives the added-risk increment;
passing `reference = NULL` assesses the total concentration. Using one
interpolation for both panels keeps the with/without-ARA comparison
apples-to-apples.

**Optional impute-first.** Both `fit_reference_model()` and `fit_target_model()`
accept an `imputation_model` (from `fit_imputation_model()`, fit on that site's
own chemistry). When supplied, missing analytes are imputed in raw concentration
space *before* modelling, so a well-sampled analyte lifts a sparsely-sampled one
into a richer spread of hydrological regimes (e.g. 100 zinc observations turn 20
copper observations into 100 copper anchors via the Cu–Zn relationship). This
requires **brms**.

---

## Reference summary statistic

The Added Risk Approach (ARA) subtracts a background "reference"
concentration before evaluating the SSD, so the AmsPAF reflects only
the *increment above* what the local community is already adapted to.
The summary statistic used to define that background is chosen by the
`summary` argument to `prepare_reference()`.

The package default is `"geom_mean"`. Rationale:

- For log-normal concentration distributions (the typical case for
  aquatic chemistry), the geometric mean is the maximum-likelihood
  central tendency.
- It uses every reference observation, making it more robust to small
  reference datasets than a fixed quantile.
- It is consistent with the PICT (pollution-induced community
  tolerance) interpretation: the community has been integrated against
  the typical exposure, not the upper tail.

Alternatives:

- `"median"` — most robust to outliers; good when reference data may be
  contaminated.
- `"arith_mean"` — uses every observation but is sensitive to outliers.
- `"p80"`, `"p90"`, `"p95"` — quantiles. Use when the assessment
  framework specifies an upper-percentile reference (e.g., some ANZG
  compliance protocols).

Set `bootstrap_ci = TRUE` to compute a 95% bootstrap CI on the chosen
summary statistic per analyte. Useful when the reference dataset is
small (n < 20 per analyte) and you want to propagate that uncertainty
visually into downstream interpretation.

## Known caveats

**Mercury and methylation.** The Hg SSD bundled with this package is for
inorganic Hg. Methylmercury (MeHg), which forms in low-redox aquatic
sediments and is the form most toxic to vertebrates, is not separately
handled — ANZG does not publish a methylmercury SSD. At sites where
reductive conditions are expected (leachate plumes, anaerobic sediment),
the inorganic Hg PAF will under-state the actual biological risk by
several orders of magnitude. This is a limitation in upstream guideline
data, not in the package.

**Mode-of-action grouping.** The current metadata places all metals into
a single `ionoregulatory` group (Concentration Addition within the
group). This is the conservative choice — it produces a higher AmsPAF
than independent action across the metals — but it is mechanistically
imprecise. Different metals act on different molecular targets (Cu/Zn on
gill Na⁺/K⁺-ATPase, Pb on calcium signalling, Cd on metallothionein,
Hg on thiol groups, As on phosphate metabolism). A more accurate
treatment would use multiple MOA groups with CA within and IA across.
This requires evidence-based assignment from the ANZG/ANZECC technical
briefs, which is tracked as a future task.

**Biological validation.** The SSDs are validated against published ANZG
DGVs (see HANDOVER.md §4) but the AmsPAF as a community-state predictor
is not yet calibrated. Until macroinvertebrate calibration is complete,
AmsPAF outputs are best-effort predictions of chronic toxicity; they
have not been empirically tied to observed community impact.

**Below-detection handling in imputation.** `fit_imputation_model()` offers
three `impute_method` options for BDL cells and cross-analyte coupling:
`"rescor_mi"` (default — full residual correlation with `mi()` imputation),
`"cens"` (proper left-censoring, no coupling), and `"cens_factor"`
(left-censoring plus a shared latent factor). The default cannot use
`cens("left")` because brms does not support censored likelihoods together with
multivariate residual correlation (`rescor = TRUE`), so it imputes BDL cells via
`mi()`. In all methods an imputed BDL value is capped at its detection limit
(`bdl_cap = TRUE`); frequent capping signals tension between the modelled
chemistry and the reported limits and should be inspected. See the *Imputing
missing and below-detection chemistry* article and the `fit_imputation_model()`
docs for the trade-offs.
