---
title: "Daily msPAF with uncertainty quantification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Daily msPAF with uncertainty quantification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4
)
```

`mspaf_daily()` can propagate uncertainty through the entire daily
interpolation pipeline, producing a credible envelope around each daily
msPAF estimate that **balloons between grab samples and pinches to
near-zero width on sampled days**.  This vignette explains the model,
shows how to use it, and describes how to interpret the output.

## Why a credible interval matters

The daily msPAF series is constructed by interpolating sparse grab
samples (typically fortnightly or monthly) onto a daily grid.  Between
grabs, concentrations are unknown: the impact model predicts a best-guess
trajectory driven by hydrology, but that prediction carries genuine
uncertainty.  Without propagating this uncertainty the point estimate can
create a false sense of precision — especially in the middle of long gaps
between samples.

The uncertainty comes from several sources:

| Source | Symbol | Description |
|--------|--------|-------------|
| Reference GAM coefficients | S1 | Posterior spread of the reference model's GAM fits (Bayesian posterior via Vp matrix) |
| Impact GAM coefficients | S2 | Same for the target model's impact-response GAMs |
| WQ-layer GAM coefficients | S3 | Same for the water quality covariate GAMs |
| State-space residual *S* | S4 | The latent residual impact *S* is a continuous-time AR(1)/Ornstein–Uhlenbeck process fitted by a Kalman smoother; the posterior variance pinches at grabs and balloons mid-gap |
| State-space WQ residual *d* | S5 | Same for the WQ-tier residual *d* |
| Anchor grab measurement error | S6 | Lognormal multiplicative error on the observed grab concentrations that anchor the bridge |
| Co-analyte grab measurement error | S7 | Same for co-analytes (pH, DOC, hardness, …) that control bioavailability normalisations |

S4 is the dominant source of uncertainty for long inter-sample gaps; S1–S3
matter most for long-range comparisons where the GAM extrapolates; S6–S7
contribute mostly at the anchor days themselves.

## The state-space smoother in one paragraph

The residual latent impact state *S* is modelled as a **continuous-time
AR(1) / Ornstein–Uhlenbeck process** and fitted to the grab-sample anchors by a
**Kalman filter + smoother** (via the **KFAS** package) over a daily grid.  The
posterior mean of the smoother *is* the centre line, and its posterior
covariance gives coherent draws — so the centre line and the credible band come
from **one** model.  The band pinches at grabs (to the grab measurement width)
and balloons mid-gap; longer gaps balloon more.

Two parameters are estimated per analyte from the anchor residuals:

* $\hat\gamma$ — the marginal (stationary) variance, the well-identified moment:
  the sample variance of the anchor *S* values.
* $\hat\theta$ — the mean-reversion rate (inverse correlation length), fitted by
  1-D maximum likelihood of the exact irregular-spacing OU likelihood, bounded so
  the correlation length stays within $[0.5\times, 10\times]$ the median anchor
  spacing.

A **fallback ladder** keeps sparse series well-behaved: analytes with fewer than
~8 anchors pin $\theta$ at the long-correlation bound (≈ a Brownian bridge);
fewer than two finite anchors, or an essentially constant series, are
"degenerate" and add no uncertainty.

**An honesty note on $\theta$.** With the ~12–30 grabs typical of routine
monitoring, $\theta$ (the correlation length) is only weakly identified
regardless of estimator — the likelihood in $\theta$ is flat because pinning the
autocorrelation decay needs many lags.  The smoother removes the bias of the old
method-of-moments estimate and makes the band *coherent*, but it cannot
manufacture information: the band is driven mainly by $\hat\gamma$, and sparse
("bridge-tier") analytes degrade gracefully to a Brownian bridge.

The daily grid is **clipped per analyte** to that analyte's grab span: no
prediction is produced before its first grab or after its last (extrapolating
the management residual beyond the data is not attempted), so coverage can be
ragged across analytes.

## Getting started

### Minimal example

```{r demo-draw, eval = FALSE}
library(hydroSense)

demo   <- leachate_demo()
target <- subset(demo, site_id == "downstream")
ref_m  <- fit_reference_model(
  subset(demo, site_id == "reference"),
  hydro = demo_hydro()
)

# 50 stochastic draws; return the 90% credible interval summary
out <- mspaf_daily(
  target,
  reference_model     = ref_m,
  interpolation       = "model",
  ndraws              = 50L,
  seed                = 42L,
  return              = "summary",
  interval            = 0.9,
  require_temperature = FALSE
)

head(out[, c("date", "mspaf", "mspaf_lower", "mspaf_upper")])
```

With `return = "summary"` (the default when `ndraws` is supplied) you get
one row per day with three columns:

* `mspaf` — the draws' own central tendency (`central = "median"` by default,
  or `"mean"`), so it is a summary *of the same posterior* as the band and
  always lies inside it
* `mspaf_lower` — lower bound of the credible interval
* `mspaf_upper` — upper bound of the credible interval

This `mspaf` is **not** the deterministic point estimate — see
[Two products: deterministic vs draws](#two-products) below.

### Including measurement error (S6 + S7)

Pass `grab_cv` to add lognormal grab-measurement error.  A scalar applies
the same CV to all analytes; a named vector sets analyte-specific CVs.

```{r grab-cv, eval = FALSE}
out_cv <- mspaf_daily(
  target,
  reference_model     = ref_m,
  interpolation       = "model",
  ndraws              = 50L,
  seed                = 42L,
  return              = "summary",
  interval            = 0.9,
  grab_cv             = 0.15,        # 15% relative analytical uncertainty
  require_temperature = FALSE
)
```

### Widening the envelope, and hydrology-driven width

`ou_scale` multiplies the marginal variance $\gamma$ (leaving $\theta$, hence the
correlation length, unchanged) — a global knob to inflate the envelope if the
default calibration feels too narrow (infrequent grabs, highly variable
analytes).

`kappa` controls **hydrology-modulated process variance**: the latent state is
allowed to wander more on high-flow days, via $q_t = q_{\text{base}}\,
e^{\kappa z_t}$ where $z_t$ is the standardised short-window hydrology feature.
This is what makes the band balloon faster across gaps that span storms (and
stay tight across quiet ones). `kappa = 0` recovers a stationary envelope; the
default `0.5` makes a ~95th-percentile flow day carry about 3× the base process
variance. It is a structural prior on volatility, not a fitted quantity.

```{r ou-scale, eval = FALSE}
out_wide <- mspaf_daily(
  target,
  reference_model = ref_m,
  interpolation   = "model",
  ndraws          = 50L,
  seed            = 42L,
  ou_scale        = 2,              # double the marginal variance (global)
  kappa           = 0.5,            # hydrology-modulated mid-gap widening
  require_temperature = FALSE
)
```

### Getting raw draws

With `return = "draws"` you get one row per (day × draw), allowing you to
build custom summaries, correlate with other variables, or feed the draws
into a downstream chronic aggregate.

```{r draws, eval = FALSE}
draws <- mspaf_daily(
  target,
  reference_model = ref_m,
  interpolation   = "model",
  ndraws          = 50L,
  seed            = 42L,
  return          = "draws",
  require_temperature = FALSE
)
# draws$draw_id runs from 1 to 50
range(draws$draw_id)
```

### Speeding things up with parallel execution

For large `ndraws` or multi-year grids, use `parallel = TRUE` with a
`future` parallel plan.  The **future.apply** package must be installed.

```{r parallel, eval = FALSE}
library(future)
plan(multisession, workers = 4)

out_par <- mspaf_daily(
  target,
  reference_model = ref_m,
  interpolation   = "model",
  ndraws          = 200L,
  seed            = 42L,
  parallel        = TRUE,
  require_temperature = FALSE
)

plan(sequential)      # restore single-threaded execution
```

Note that parallel draws use an independent RNG stream
(L'Ecuyer-CMRG via `future.apply`), so they will differ numerically from a
sequential run with the same seed, but are themselves reproducible.

## Plotting the credible envelope

```{r plot, eval = FALSE}
library(ggplot2)

ggplot(out, aes(date)) +
  geom_ribbon(aes(ymin = mspaf_lower, ymax = mspaf_upper),
              fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = mspaf), colour = "steelblue", linewidth = 0.8) +
  scale_y_continuous(labels = scales::label_percent(scale = 1)) +
  labs(
    title    = "Daily msPAF with 90% credible interval",
    subtitle = "Ribbon balloons mid-gap, pinches at grab-sample dates",
    x = NULL, y = "% species affected"
  ) +
  theme_minimal()
```

## Interpreting the width of the envelope

**Narrow ribbon at a grab date**: the smoother is pinned to the observed
concentration (to within the grab measurement width).  Remaining width comes
from GAM coefficient uncertainty (S1–S3) and measurement error (S6–S7).

**Wide ribbon mid-gap**: the smoother posterior has spread away from the
anchors.  The width grows with the gap length, the estimated $\gamma$
(residual-state marginal variance), and — through `kappa` — the hydrology over
the gap, so a gap spanning a storm balloons faster than a quiet one of equal
length.  A site with infrequent, irregular sampling shows a wider envelope than
one with fortnightly grabs.

**Asymmetric ribbon**: the msPAF response to concentration is nonlinear
(SSD-based) and the impact is floored at zero, so the ribbon is skewed even
though the underlying concentration draws are symmetric.  The *direction* of the
skew is data-dependent: symmetric concentration draws map through the SSD
curvature, which can make the msPAF band wider above *or* below the centre line
depending on where on the SSD the site sits.

**Envelope does not cover the centre line**: this should never happen in
practice, but if your `ndraws` is very small (e.g., 5) and the interval is
wide (e.g., 0.99), quantile estimates can be noisy.  Increase `ndraws`.

## Parameter interpretation table

| Parameter | What it controls | Typical values |
|-----------|-----------------|----------------|
| `ndraws` | Number of Monte Carlo draws; more → smoother CI | 50–200 for exploratory work; 500+ for publication |
| `interval` | Nominal coverage of the credible interval | 0.9 (default), 0.95, 0.80 |
| `grab_cv` | Relative analytical uncertainty on grab concentrations | 0.10–0.20 for routine ICP-MS; 0.05 for reference standards |
| `ou_scale` | Global marginal-variance ($\gamma$) inflation; > 1 widens the whole envelope | 1 (default); try 2–4 for sparse sampling |
| `kappa` | Hydrology-modulated mid-gap widening ($q_t = q_{\text{base}}e^{\kappa z_t}$) | 0.5 (default); 0 = stationary |
| `seed` | RNG seed for reproducibility | Any integer |
| `parallel` | Use `future.apply` for concurrent draws | `FALSE` (default); `TRUE` with a `future::plan()` |

## Two products: deterministic vs draws {#two-products}

`mspaf_daily()` gives you **one of two distinct products**, selected by
`ndraws`. They answer different questions and are not expected to coincide.

**Deterministic** (`ndraws = NULL`, the default).
A single grabs-exact daily line: the residual smoother is pinned to the measured
grab samples, so the curve threads the reported lab values.

* *Pro* — fast, fully reproducible, the best single-number estimate; an ideal
  sanity check against the draws.
* *Con* — carries no uncertainty, and treats the grabs as exact (it over-fits
  noisy lab values).

**Draws** (`ndraws > 0`).
A Monte-Carlo posterior that propagates trend (GAM), between-grab (OU bridge)
and — when `grab_cv` is set — grab measurement uncertainty. `return = "summary"`
reports the draws' central tendency (`central`) plus a credible band;
`return = "draws"` returns the raw per-draw paths.

* *Pro* — honest uncertainty quantification; the summary `mspaf` is the band's
  own centre (it always lies inside `[mspaf_lower, mspaf_upper]`).
* *Con* — the central estimate now depends on `ndraws`/`seed`, and by Jensen's
  inequality on the bounded msPAF index it generally sits *above* the
  deterministic line; adding `ndraws` therefore changes the point estimate.

A natural workflow is to compute the deterministic line (cheap) **and** the
draws, then overlay them: the deterministic is your best guess, the band shows
the uncertainty around it. If the deterministic strays well outside the band,
that is a useful diagnostic, not a contradiction — the two summarise different
posteriors.

> **Grab measurement error (`grab_cv`).** Until per-sample lab uncertainties are
> available, leave `grab_cv = NULL` (the default): a fabricated CV injects
> measurement spread that is not real. When real uncertainties land, the draws
> should use them (widening the band, and pinching it toward the measurement
> width at grab dates), while the deterministic line continues to thread the
> reported values.

## Connection to the chronic pipeline

The `return = "draws"` output can be passed into `time_weighted_aggregate()`
to propagate daily uncertainty through to the chronic msPAF (via the
draw-carrier contract).  The chronic draws can then be summarised with
`summarise_draws()`.

```{r chronic, eval = FALSE}
draws_daily  <- mspaf_daily(..., return = "draws")
draws_mspaf <- dplyr::filter(draws_daily, FALSE)  # placeholder
# See vignette("chronic-mspaf-interpretation") for the chronic pathway
```

See the *Uncertainty propagation* section of
`vignette("chronic-mspaf-interpretation")` for the end-to-end workflow.

## See also

* [mspaf_daily()] — full parameter reference
* [fit_reference_model()] — fitting the reference/impact model
* [summarise_draws()] — collapse draws to median + credible interval
* `vignette("chronic-mspaf-interpretation")` — interpreting chronic msPAF
