---
title: "Daily AmsPAF with uncertainty quantification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Daily AmsPAF with uncertainty quantification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4
)
```

`amspaf_daily()` can propagate uncertainty through the entire daily
interpolation pipeline, producing a credible envelope around each daily
AmsPAF estimate that **balloons between grab samples and pinches to
near-zero width on sampled days**.  This vignette explains the model,
shows how to use it, and describes how to interpret the output.

## Why a credible interval matters

The daily AmsPAF series is constructed by interpolating sparse grab
samples (typically fortnightly or monthly) onto a daily grid.  Between
grabs, concentrations are unknown: the impact model predicts a best-guess
trajectory driven by hydrology, but that prediction carries genuine
uncertainty.  Without propagating this uncertainty the point estimate can
create a false sense of precision — especially in the middle of long gaps
between samples.

The uncertainty comes from several sources:

| Source | Symbol | Description |
|--------|--------|-------------|
| Reference GAM coefficients | S1 | Posterior spread of the reference model's GAM fits (Bayesian posterior via Vp matrix) |
| Impact GAM coefficients | S2 | Same for the target model's impact-response GAMs |
| WQ-layer GAM coefficients | S3 | Same for the water quality covariate GAMs |
| OU bridge on residual state *S* | S4 | The latent residual impact state *S* between grab anchors follows an Ornstein–Uhlenbeck bridge: variance 0 at anchors, maximal mid-gap |
| OU bridge on WQ score *d* | S5 | Same for the WQ-score prediction in leading and trailing zones |
| Anchor grab measurement error | S6 | Lognormal multiplicative error on the observed grab concentrations that anchor the bridge |
| Co-analyte grab measurement error | S7 | Same for co-analytes (pH, DOC, hardness, …) that control bioavailability normalisations |

S4 is the dominant source of uncertainty for long inter-sample gaps; S1–S3
matter most for long-range comparisons where the GAM extrapolates; S6–S7
contribute mostly at the anchor days themselves.

## The OU bridge in one paragraph

Between two consecutive grab dates (anchors), the residual latent impact
state *S* is modelled as an **Ornstein–Uhlenbeck bridge** — a continuous-time
AR(1) process conditioned to equal zero at both ends.  The marginal variance
of the bridge at time *t* within a gap of length *L* is

$$\mathrm{Var}[S(t)] = \gamma \frac{(1-u^2)(1-v^2)}{1-s^2}$$

where $u = e^{-\theta t}$, $v = e^{-\theta(L-t)}$, $s = e^{-\theta L}$, and
$\gamma = \sigma^2/(2\theta)$ is the marginal variance of the stationary OU
process.  As $\theta \to 0$ this converges exactly to the Brownian bridge
$\sigma^2 t(L-t)/L$.

The parameters $\theta$ (mean-reversion rate, 1/day) and $\gamma$ are
estimated by method of moments from the observed residual-state values at
the grab-sample anchor dates:

* $\hat\gamma$ = sample variance of the anchor *S* values
* $\hat\sigma^2$ = median quadratic variation $(\Delta S)^2 / \Delta t$
  over consecutive anchor pairs
* $\hat\theta = \hat\sigma^2 / (2\hat\gamma)$

When fewer than two finite anchors are available, or when the anchor
series is essentially constant, the estimator returns a "degenerate"
flag and the bridge collapses to zero — no uncertainty is added.

## Getting started

### Minimal example

```{r demo-draw, eval = FALSE}
library(leachatetools)

demo   <- leachate_demo()
target <- subset(demo, site_id == "downstream")
ref_m  <- fit_reference_model(
  subset(demo, site_id == "reference"),
  hydro = demo_hydro()
)

# 50 stochastic draws; return the 90% credible interval summary
out <- amspaf_daily(
  target,
  reference_model     = ref_m,
  interpolation       = "model",
  ndraws              = 50L,
  seed                = 42L,
  return              = "summary",
  interval            = 0.9,
  require_temperature = FALSE
)

head(out[, c("date", "amspaf", "amspaf_lower", "amspaf_upper")])
```

With `return = "summary"` (the default when `ndraws` is supplied) you get
one row per day with three columns:

* `amspaf` — deterministic centre-line (heuristic-blend point prediction,
  identical to the output of a non-draws run with `interpolation = "model"`)
* `amspaf_lower` — lower bound of the credible interval
* `amspaf_upper` — upper bound of the credible interval

### Including measurement error (S6 + S7)

Pass `grab_cv` to add lognormal grab-measurement error.  A scalar applies
the same CV to all analytes; a named vector sets analyte-specific CVs.

```{r grab-cv, eval = FALSE}
out_cv <- amspaf_daily(
  target,
  reference_model     = ref_m,
  interpolation       = "model",
  ndraws              = 50L,
  seed                = 42L,
  return              = "summary",
  interval            = 0.9,
  grab_cv             = 0.15,        # 15% relative analytical uncertainty
  require_temperature = FALSE
)
```

### Widening the bridge envelope

The `ou_scale` argument multiplies $\sigma^2$ and $\gamma$ (without
changing $\theta$, so the correlation length is preserved).  Use it to
inflate the OU envelope if the default calibration feels too narrow for
your data — for example when grab samples are infrequent or analyte
concentrations are highly variable.

```{r ou-scale, eval = FALSE}
out_wide <- amspaf_daily(
  target,
  reference_model = ref_m,
  interpolation   = "model",
  ndraws          = 50L,
  seed            = 42L,
  ou_scale        = 2,              # double the bridge variance
  require_temperature = FALSE
)
```

### Getting raw draws

With `return = "draws"` you get one row per (day × draw), allowing you to
build custom summaries, correlate with other variables, or feed the draws
into a downstream chronic aggregate.

```{r draws, eval = FALSE}
draws <- amspaf_daily(
  target,
  reference_model = ref_m,
  interpolation   = "model",
  ndraws          = 50L,
  seed            = 42L,
  return          = "draws",
  require_temperature = FALSE
)
# draws$draw_id runs from 1 to 50
range(draws$draw_id)
```

### Speeding things up with parallel execution

For large `ndraws` or multi-year grids, use `parallel = TRUE` with a
`future` parallel plan.  The **future.apply** package must be installed.

```{r parallel, eval = FALSE}
library(future)
plan(multisession, workers = 4)

out_par <- amspaf_daily(
  target,
  reference_model = ref_m,
  interpolation   = "model",
  ndraws          = 200L,
  seed            = 42L,
  parallel        = TRUE,
  require_temperature = FALSE
)

plan(sequential)      # restore single-threaded execution
```

Note that parallel draws use an independent RNG stream
(L'Ecuyer-CMRG via `future.apply`), so they will differ numerically from a
sequential run with the same seed, but are themselves reproducible.

## Plotting the credible envelope

```{r plot, eval = FALSE}
library(ggplot2)

ggplot(out, aes(date)) +
  geom_ribbon(aes(ymin = amspaf_lower, ymax = amspaf_upper),
              fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = amspaf), colour = "steelblue", linewidth = 0.8) +
  scale_y_continuous(labels = scales::label_percent(scale = 1)) +
  labs(
    title    = "Daily AmsPAF with 90% credible interval",
    subtitle = "Ribbon balloons mid-gap, pinches at grab-sample dates",
    x = NULL, y = "% species affected"
  ) +
  theme_minimal()
```

## Interpreting the width of the envelope

**Narrow ribbon at a grab date**: the bridge is pinned to the observed
concentration.  Remaining width comes from GAM coefficient uncertainty
(S1–S3) and measurement error (S6–S7).

**Wide ribbon mid-gap**: the OU bridge has spread away from the anchors.
The width is controlled by the estimated $\gamma$ (residual-state marginal
variance from the anchor series) and the gap length.  A site with
infrequent, irregular sampling will show a wider envelope than one with
fortnightly grabs.

**Asymmetric ribbon**: the response of AmsPAF to concentration is
nonlinear (SSD-based), so the ribbon can be skewed even when the underlying
concentration draws are symmetric.

**Envelope does not cover the centre line**: this should never happen in
practice, but if your `ndraws` is very small (e.g., 5) and the interval is
wide (e.g., 0.99), quantile estimates can be noisy.  Increase `ndraws`.

## Parameter interpretation table

| Parameter | What it controls | Typical values |
|-----------|-----------------|----------------|
| `ndraws` | Number of Monte Carlo draws; more → smoother CI | 50–200 for exploratory work; 500+ for publication |
| `interval` | Nominal coverage of the credible interval | 0.9 (default), 0.95, 0.80 |
| `grab_cv` | Relative analytical uncertainty on grab concentrations | 0.10–0.20 for routine ICP-MS; 0.05 for reference standards |
| `ou_scale` | Envelope inflation factor; > 1 widens mid-gap spread | 1 (default); try 2–4 for sparse sampling |
| `seed` | RNG seed for reproducibility | Any integer |
| `parallel` | Use `future.apply` for concurrent draws | `FALSE` (default); `TRUE` with a `future::plan()` |

## How the centre line is computed

The `amspaf` column in `return = "summary"` is the **deterministic
heuristic-blend point prediction**, which is identical to running
`amspaf_daily()` without `ndraws`.  It is *not* the median of the draws.
This means the centre line:

* is computationally cheap (no randomness),
* is exactly reproducible regardless of `ndraws` or `seed`,
* is identical to the pre-uncertainty-propagation output, so adding `ndraws`
  to an existing call does not change the point estimate.

The lower and upper bounds are empirical quantiles of the stochastic draws.

## Connection to the chronic pipeline

The `return = "draws"` output can be passed into `time_weighted_aggregate()`
to propagate daily uncertainty through to the chronic AmsPAF (via the
draw-carrier contract).  The chronic draws can then be summarised with
`summarise_draws()`.

```{r chronic, eval = FALSE}
draws_daily  <- amspaf_daily(..., return = "draws")
draws_amspaf <- dplyr::filter(draws_daily, FALSE)  # placeholder
# See vignette("chronic-amspaf-interpretation") for the chronic pathway
```

See the *Uncertainty propagation* section of
`vignette("chronic-amspaf-interpretation")` for the end-to-end workflow.

## See also

* [amspaf_daily()] — full parameter reference
* [fit_reference_model()] — fitting the reference/impact model
* [summarise_draws()] — collapse draws to median + credible interval
* `vignette("chronic-amspaf-interpretation")` — interpreting chronic AmsPAF
