---
title: "Imputing missing and below-detection chemistry"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Imputing missing and below-detection chemistry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE
)
library(leachatetools)
```

## The problem

Routine monitoring chemistry is rarely complete. Two kinds of gaps dominate:

- **Below detection limit (BDL).** The analyte was measured but the result
  fell under the laboratory's reporting limit. All we know is that the true
  value lies somewhere in `[0, DL]`. Substituting `DL`, `DL/2`, or `0` —
  the usual quick fixes — biases every downstream summary, and the direction
  of the bias depends on which fix you picked.
- **Missing.** The analyte was not measured at that sample at all, so there
  is no value and no detection limit.

Both gaps matter for mixture toxicity: `add_amspaf()` sums the potentially
affected fraction across analytes, so a sample where Cu was never measured
and Zn came back BDL is not automatically "clean" — it is *under-determined*.
The imputation tools fill these gaps with statistically defensible estimates
that borrow strength from the analytes that *were* measured.

## The approach in brief

leachatetools imputes chemistry with a **Bayesian multivariate model** fitted
once on your monitoring history and then applied to new samples. The design
has three ideas at its core:

1. **A shared chemistry summary as predictors.** All the routinely measured
   water-quality variables — major ions, pH, EC, NH<sub>3</sub>-N, DOC,
   nutrients, redox indicators — are compressed into a handful of principal
   components with `nipals::nipals()`, which tolerates within-sample missing
   cells natively. These PC scores become the predictors, so a missing metal
   is estimated from the *whole* measured chemical context, not one or two
   surrogate analytes.

2. **Cross-analyte coupling.** Metals are modelled jointly. In a leachate- or
   AMD-impacted aquifer, conservative tracers move ahead of redox-controlled
   metals, so a post-pulse sample can look near-baseline on its major ions
   while Cu/Pb/Zn/Mn stay co-elevated. That co-elevation is residual
   correlation with no predictor driving it — capturing it lets an *observed*
   metal at a sample inform the *missing* ones. By default this is a full
   residual correlation matrix (`rescor = TRUE`); see
   *Choosing the imputation method* for the alternatives.

3. **Hurdles, so silence is not mistaken for absence.** A sample is only
   imputed for a group if it carries at least one analyte from that group —
   one metal for the metals model, one of {DOC, TOC, BOD, COD, cBOD} for the
   organics model. A sample with no metals recorded is left alone, because a
   leachate metal pulse may simply not have reached that location yet, and
   inventing values there would manufacture an impact that the data do not
   support.

The work is split across three functions:

| Function | Role | Engine |
|----------|------|--------|
| `fit_imputation_model()` | Fit the metals/organics model(s) once on training data | `brms` (Stan) |
| `impute_chemistry()` | Fill BDL and missing metals/organics on new data | `brms` posterior |
| `impute_coanalytes()` | Fill missing normalisation co-analytes (DOC, Ca, Mg, hardness) | `mgcv::gam` |

## The data contract

Every function takes the same long-format frame used throughout the package:
one row per sample × analyte, with `sample_id`, `site_id`, `datetime`,
`analyte`, `value`, and `detected`. A BDL result is a row where `detected`
is `FALSE` and `value` holds the detection limit; a missing result is simply
the absence of a row.

```{r contract}
chem <- tibble::tribble(
  ~sample_id, ~analyte, ~value, ~detected,
  "s1",       "pH",       7.2,   TRUE,
  "s1",       "EC",      410,    TRUE,
  "s1",       "Cu",        3.1,  TRUE,
  "s1",       "Zn",        0.5,  FALSE,   # BDL: value = detection limit
  "s1",       "DOC",       2.4,  TRUE
  # Ni simply has no row for s1 -> "missing"
)
chem
```

`pH` and `EC` are **required variables**: a sample missing either is dropped
before fitting, because they anchor the chemistry PCA. When a required
variable is itself BDL, its detection-limit value is used as a conservative
stand-in and the sample is kept.

## Fitting the model

`fit_imputation_model()` discovers the analyte groups automatically: anything
matching the built-in metals list becomes the metals model, the remaining
non-excluded concentration analytes become the organics model, and
microbiological counts and qualitative descriptors are excluded outright.
Fitting runs MCMC through Stan, so it takes minutes, not seconds — do it once
and reuse the result.

```{r fit, eval = FALSE}
model <- fit_imputation_model(
  monitoring_long,
  required_vars = c("pH", "EC"),
  save_dir      = "models"   # optional: persist the fit as a .qs file
)

model
#> <imputation_model>  fitted 2026-06-03 | 248 samples | 14 PCA vars | 3 PCs (78% var)
#>   method:   rescor_mi
#>   metals:   8 analytes
#>   organics: 2 analytes
```

The PCA axis count is chosen to reach a target cumulative variance
(`min_var_explained`, default 75%) up to `max_pcs` (default 6), with a floor
of two axes. The console reports which variables entered the PCA, how many
axes were retained, and the analytes in each group, so you can confirm the
model saw what you expected.

A metal or organic is only modelled if it is *detected* in at least
`min_target_detect_freq` of samples (default 5%). Near- or all-below-detection
analytes carry no signal to impute from and would otherwise inflate the model
— on a panel with a hundred mostly-undetected organics this filter is what
keeps the fit tractable.

## Choosing the imputation method

`fit_imputation_model(impute_method =)` controls how below-detection cells and
cross-analyte coupling are handled. All three share the same PCA predictors;
`"rescor_mi"` and `"cens"` fit a wide multivariate model, while `"cens_factor"`
fits a single long-format model with a shared per-sample factor:

| `impute_method` | BDL handling | Coupling | Notes |
|-----------------|--------------|----------|-------|
| `"rescor_mi"` *(default)* | `mi()` (imputed) + DL cap | full residual-correlation matrix | most accurate recovery; posterior **draws are memory-heavy** |
| `"cens"` | `cens("left")` (proper censoring) | none | fastest; cleanest convergence; no borrowing across metals |
| `"cens_factor"` | `cens("left")` | shared per-sample latent factor | proper censoring **and** coupling; calibrated, cheap uncertainty |

The trade-off is accuracy versus the cost of uncertainty. `brms` cannot combine
a full residual correlation (`rescor = TRUE`) with proper left-censoring, so the
default pairs residual correlation with `mi()` imputation and a post-hoc cap.
A hold-out benchmark on real data found `"rescor_mi"` recovers masked metals most
accurately (the full residual-correlation matrix is the strongest coupling),
which is why it is the default — but its posterior **draws** are expensive and
can exhaust memory.

`"cens"` and `"cens_factor"` give well-calibrated intervals cheaply. They differ
in coupling: `"cens"` treats the metals independently, while `"cens_factor"`
adds a single per-sample latent factor shared across all analytes — fitted in
long format, so the factor is well-identified (each sample contributes several
analyte observations) and genuinely lets an observed metal inform the
unobserved/BDL ones at that sample. In the B.S01 hold-out this coupling places
`"cens_factor"` between `"cens"` and `"rescor_mi"` on accuracy (RMSE 0.43 vs
0.53 and 0.25) while converging best of the three (R̂ 1.006). Prefer a cens
method when you need
uncertainty (`return = "draws"`) or hit memory limits — `"cens_factor"` when you
also want cross-analyte borrowing, `"cens"` when you want the simplest, fastest
fit; prefer the default `"rescor_mi"` for the most accurate point estimates.

```{r method, eval = FALSE}
# Proper censoring + cross-analyte coupling, with cheap calibrated uncertainty:
model <- fit_imputation_model(monitoring_long, impute_method = "cens_factor")
```

## Imputing new chemistry

`impute_chemistry()` applies the fitted model, returning the input frame with
BDL and missing metals/organics cells replaced by posterior estimates and two
new columns:

- `imputed` — `TRUE` for any cell that was filled.
- `imputed_kind` — `"observed"`, `"censored_left"` (was BDL), or `"missing"`
  (had no row).

```{r impute, eval = FALSE}
filled <- impute_chemistry(monitoring_long, model)

# Inspect what was filled, and how
dplyr::count(filled, analyte, imputed_kind)
```

By default the result is a **posterior mean per cell** (`return = "point"`).
For uncertainty propagation, request the full posterior with
`return = "draws"` to get one row per sample × analyte × draw — feed those
draws through `add_amspaf()` to obtain a posterior distribution of mixture
PAF rather than a point estimate. Draws can be memory-heavy (especially for
`"rescor_mi"`); `ndraws =` subsamples the posterior and `batch_size =` predicts
the samples in batches, both of which bound peak memory.

### The detection-limit cap

An imputed below-detection value must never exceed its detection limit. The
posterior prediction is not itself constrained below the DL, so
`impute_chemistry()` **caps** any imputed BDL cell at its original detection
limit (`bdl_cap = TRUE`, the default) and warns when the cap fires, with a
per-analyte breakdown (how many cells, and the worst exceedance ratio) so the
activations are auditable rather than just an aggregate count. Frequent cap
firing signals tension between the modelled chemistry and the reported limits.

For a full per-cell audit, the result carries a `"bdl_cap_summary"` attribute —
one row per capped (`sample_id`, `analyte`) cell with its detection limit, the
uncapped estimate, and the exceedance ratio. Retrieve it with
`bdl_cap_summary()` on the frame as returned (plain attributes are dropped by
most dplyr verbs, so read it before further wrangling):

```{r capsummary, eval = FALSE}
filled <- impute_chemistry(monitoring_long, model)
bdl_cap_summary(filled)   # NULL (with a message) if nothing was capped
```

## Imputing normalisation co-analytes

The bioavailability normalisations in `add_amspaf()` need co-analytes —
DOC, Ca, Mg, and hardness — to convert field concentrations to the SSD index
condition (see the *Analyte normalisation* article). When one of these is
missing, `impute_coanalytes()` fills it with a fast, deterministic
log-Gaussian GAM (`mgcv::gam`) on the *same* PC scores the metals model used,
so the co-analyte estimate is conditioned on the same chemical context.

```{r coanalytes, eval = FALSE}
filled <- filled |>
  impute_coanalytes(model)        # fills DOC / Ca / Mg / hardness where absent
```

Two deliberate boundaries:

- It only fills analytes that are **entirely absent** for a sample; existing
  BDL observations are left untouched.
- Imputed co-analyte values are **never** fed back into the metals/organics
  model — that model was already fitted on measured values. Co-analyte
  imputation is a downstream convenience for normalisation, not part of the
  Bayesian fit.

## Where it fits in the pipeline

The imputation steps slot in before mixture toxicity and temporal
aggregation:

```{r pipeline, eval = FALSE}
model  <- fit_imputation_model(monitoring_long)

result <- monitoring_long |>
  impute_chemistry(model) |>     # BDL + missing metals/organics
  impute_coanalytes(model) |>    # missing DOC / Ca / Mg / hardness
  add_amspaf(reference = prep_ref) |>
  time_weighted_aggregate()
```

Fit once on the longest, richest history you have; impute repeatedly as new
data arrive.

## When not to impute

Imputation borrows strength across analytes — it does not conjure information
that the data never contained. A few situations call for restraint:

- **Sparse or idiosyncratic chemistry.** If a sample shares little measured
  context with the training data, its PC scores are poorly determined and the
  posterior is wide. The hurdles guard the worst cases, but a sample that
  barely clears a hurdle still deserves a look.
- **Near-all-below-detection organics.** In landfill/WMF leachate most organic
  analytes are below detection in almost every sample, so there is no signal to
  borrow from and no reliable covariate to anchor them — DOC fixes the bulk
  carbon, not a congener-specific ratio. The package deliberately does **not**
  impute these: the DOC-like hurdle skips the organics model without a carbon
  covariate, and `min_target_detect_freq` (default 5%) drops any analyte
  detected too rarely to model. Such organics are left as reported
  non-detections rather than given manufactured values; treat them as a
  censored "≤ detection limit" reading in any downstream summary.
- **Frequent BDL-cap warnings for an analyte.** This signals tension between
  the modelled chemistry and the reported detection limits; treat those cells
  as flagged rather than final.
- **Reporting raw detections.** Imputed values are estimates. Keep the
  `imputed` / `imputed_kind` columns with your results so that a reader can
  always separate measured from modelled, and present the posterior draws
  where the downstream decision is sensitive to the gaps.

## Installation note

The Bayesian step is the only part of leachatetools that needs **brms**, so
brms is an optional dependency rather than a hard requirement — the LMF and
AmsPAF tools install and run without it. To use `fit_imputation_model()` and
`impute_chemistry()`, install brms and a working Stan engine once:

```r
install.packages("brms")
```

If the install alone is not enough, follow the short Stan toolchain setup at
<https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started>.
`impute_coanalytes()` uses `mgcv` (already a hard dependency) and does **not**
require brms or Stan.
