---
title: "Detecting leachate with the Leachate Mixing Fraction (LMF)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Detecting leachate with the Leachate Mixing Fraction (LMF)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE
)
library(leachatetools)
```

## What the LMF answers

`add_lmf()` asks a single question of each water sample:

> *What fraction of this sample's chemistry can be explained as a mixture of
> local reference water and landfill leachate?*

The answer — the **leachate mixing fraction (LMF)** — is reported as a
percentage: `0` means the sample is chemically indistinguishable from local
reference water, `100` means it matches the leachate end-member. It is a
*source-detection* index, deliberately separate from toxicity: a sample can
carry a clear leachate signature (high LMF) while still being well below
guideline values, and vice versa.

LMF is built on the classic **two-component end-member mixing** model used in
catchment hydrochemistry (Christophersen & Hooper, 1992). Leachate-impacted
freshwater is treated as a blend of two sources, and the major-ion signature is
used to estimate the blend ratio.

## The method in brief

For each sample, `add_lmf()`:

1. **Converts to milliequivalents.** Concentrations are converted to meq/L with
   [to_meq()] so that ions combine on a charge-equivalent basis. Transforming
   species are collapsed to conserved totals — the nitrogen species
   (NH<sub>3</sub>-N, NO<sub>3</sub>-N, NO<sub>2</sub>-N) to *total N*, and the
   carbonate species to *total alkalinity* — so that redox or speciation shifts
   between source and sample do not break the mixing assumption.

2. **Builds two end-members.** The **reference** end-member is the mean and
   standard deviation of each ion in local background water. The **leachate**
   end-member is anchored to chloride (a conservative tracer) and expressed as
   Cl-scaled ratios.

3. **Computes a per-ion mixing fraction** for each measured ion *i*:
   $$ f_i = \frac{x_i - R_i}{L_i - R_i} $$
   where $x_i$ is the sample, $R_i$ the reference, and $L_i$ the leachate
   end-member.

4. **Combines the ions** into one estimate by an inverse-variance weighted
   mean, so ions measured precisely and with a large leachate–reference gradient
   count for more. An **informativeness** screen (the reference scatter relative
   to the leachate–reference gap) admits a sample only if enough
   high-information ions are present.

5. **Down-weights disagreeing ions** with a robust (Huber) reweighting pass, so
   a single ion behaving non-conservatively (sulfate reduction, an extra source)
   does not drag the estimate. A χ² diagnostic flags overall ion disagreement.

## A worked example

The package ships a small synthetic dataset, `leachate_demo`, with three sites:
an impacted `downstream` site, a clean `reference` site, and the `leachate`
end-member. (It is fictional — generated from a mixing model so the example is
reproducible.)

```{r data}
demo <- leachate_demo()
unique(demo$site_id)
head(subset(demo, site_id == "downstream"), 4)
```

Supplying the leachate and reference chemistry directly uses the
caller-provided end-member path (no site-matching infrastructure required):

```{r run-lmf}
out <- add_lmf(
  df             = subset(demo, site_id == "downstream"),
  leachate_data  = subset(demo, site_id == "leachate"),
  reference_data = subset(demo, site_id == "reference")
)

lmf <- subset(out, analyte == "LMF",
              c("sample_id", "value", "sigma_lmf", "n_ions_used", "chi2_per_df"))
lmf
```

The downstream samples return an LMF of about 15% — consistent with how the
demo data were constructed (a 15% leachate blend). `sigma_lmf` is the
uncertainty on the estimate (percentage points), `n_ions_used` the number of
ions that contributed, and `chi2_per_df` the ion-agreement diagnostic (values
near 1 indicate the ions tell a consistent mixing story).

## Interpreting the result

LMF is a continuous index: the raw percentage is the primary output. The
following tier breaks are a suggested starting point for interpretation:

| Tier | LMF (%) | Suggested interpretation |
|------|---------|--------------------------|
| 1 | ≤ 1 | Background — indistinguishable from local reference |
| 2 | 1–5 | Trace impact — detectable signature, monitor trend |
| 3 | 5–20 | Significant impact — clear signature, investigate |
| 4 | > 20 | Severe impact — urgent investigation |

> **Note:** These breaks are **starting points**, not regulatory thresholds.
> Classification decisions should be made by the reporting system or analyst,
> not imposed automatically. The tier 1/2 boundary in particular should be
> reviewed against the LMF distribution of your own reference samples after
> deployment — ideally it should sit near the 95th percentile of LMF on
> known-clean water.

## When LMF returns `NA`

A sample is not scored (its `lmf_reason` explains why) when it lacks enough
high-information ions, when the estimate is too imprecise (`sigma_lmf` above
the `max_sigma_lsi` threshold), or when no ion has a usable
leachate–reference gradient. These are honest non-answers: the major-ion panel
measured at that sample simply cannot distinguish the two sources.

## References

- Christophersen N, Hooper RP (1992). Multivariate analysis of stream water
  chemical data. *Water Resources Research* 28(1):99–107.
- Christensen JB et al. (2001). Biogeochemistry of landfill leachate plumes.
  *Applied Geochemistry* 16(7–8):659–718.
- Kjeldsen P et al. (2002). Present and long-term composition of MSW landfill
  leachate. *Critical Reviews in Environmental Science and Technology* 32(4).
